---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## **Loading and preprocessing the data**  

Loading and seeing data

```{r cache=TRUE}
library(readr)
library(dplyr)
data <- read_csv("activity.csv")
head(data)
str(data)
```

Format date and interval variables

```{r cache=TRUE}

data$date <- as.Date(data$date, format = "%Y-%m-%d")
# data$interval <- factor(data$interval)
```


## **What is mean total number of steps taken per day?**  

Steps per day

```{r cache=TRUE}
steps <- data %>% 
  group_by(date) %>% 
  summarise(steps = sum(steps, na.rm = TRUE))

head(steps)
```

Histogram of steps per day

```{r cache=TRUE}
hist(steps$steps, main = "Total Steps Per Day", xlab = "Steps", 
     col = "dark blue")
```
  
Mean steps by day

```{r cache=TRUE}
mean(steps$steps)
```

Median of steps by day

```{r cache=TRUE}
median(steps$steps)
```


## **What is the average daily activity pattern?**

Average steps across all days
```{r cache=TRUE}
steps_interval <- data %>% 
  group_by(interval) %>% 
  summarise(meansteps = mean(steps, na.rm = TRUE))

head(steps_interval)
```

Plot of average steps per day by interval

```{r cache=TRUE}
plot(steps_interval$meansteps ~ steps_interval$interval, col = "dark blue",
     type = "l", main = "Mean Steps By Time Interval", xlab = "5min Interval",
     ylab = "Mean Steps")

```

Interval containing the most steps on average
```{r cache=TRUE}
most_steps <- steps_interval$interval[which.max(steps_interval$meansteps)]
print(most_steps)
```

Average steps for the previous interval
```{r cache=TRUE}
average_most <- max(steps_interval$meansteps)
print(average_most)
```


## **Imputing missing values**

Total number of NA's
```{r cache=TRUE}
print(sum(is.na(data$steps)))
```

NA's Strategy:  
Replacing steps NA's with mean steps for the interval.  
  
Creating a new dataset without NA values
```{r cache=TRUE}
data_nona <- data

for (i in 1:nrow(data)) {
  if(is.na(data$steps[i])) {
    data_nona$steps[i] <- steps_interval$meansteps[data_nona$interval[i] == 
                                                     steps_interval$interval]
  }
}

head(data_nona)
```

New histogram of total steps taken per day

```{r cache=TRUE}
steps_day2 <- data_nona %>% 
  group_by(date) %>% 
  summarise(steps = sum(steps))

hist(steps_day2$steps, main = "New Total Steps Per Day", xlab = "Steps", 
     col = "dark blue")
```

New mean steps by day

```{r cache=TRUE}
mean(steps_day2$steps)
```

New median of steps by day

```{r cache=TRUE}
median(steps_day2$steps)
```

The new average is different because on the original NA values were removed. After filling NA
values, the average increased.


## **Are there differences in activity patterns between weekdays and weekends?**

Creating factor variable for weekday and weekend and filter each activity

```{r cache=TRUE}
data2 <- data_nona
data2$date <- as.Date(data2$date)
data2$day <- ifelse(weekdays(data2$date) %in% c("Saturday", "Sunday"), 
                    "weekend", "weekday")

weekday <- filter(data2, day == "weekday") %>% 
  group_by(interval) %>% 
  summarise(steps = mean(steps)) %>% 
  mutate(day = as.factor("weekday"))

weekend <- filter(data2, day == "weekend") %>% 
  group_by(interval) %>% 
  summarise(steps = mean(steps)) %>% 
  mutate(day = as.factor("weekend"))

```

Creating new dataset

```{r cache=TRUE}
data_days <- rbind(weekday, weekend)
data_days$day <- as.factor(data_days$day)

```

Plotting weekday and weekend activity

```{r cache=TRUE}
library(ggplot2)
ggplot(data_days, aes(interval, steps)) +
  geom_line() +
  facet_grid(day~.) +
  ggtitle("Average steps: Weekday vs Weekend") +
  xlab("Interval") +
  ylab("Steps")
```
  
There are little differences in steps patterns by interval on weekdays and weekend.
